---
title: "Typical message making tasks and challenges"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = TRUE,
  error = TRUE
)
```

We need to emit:

* Informational messages
* Error messages

<aside>
I'm ignoring warnings, because I use them so seldom, but a full treatment should cover them. Our lifecycle policies, in particular, lean heavily on warnings.
</aside>

What do these messages tend to look like, in the wild, using base tools?

## `base::message()` and `base::stop()`

First, let's review the signatures of `message()` and `stop()`.

```{r eval = FALSE}
message(..., domain = NULL, appendLF = TRUE)
stop(..., call. = TRUE, domain = NULL)
```

Both make make a message from the loose parts that come via `...`, which is documented as:

> zero or more objects which can be coerced to character (and which are pasted together with no separator) 

Here's a tour of `message()` calls sprinkled around tidyverse / r-lib, with me pointing out small, recurring challenges.
I'll often have to make some fake objects, so we can actually run the code.

This is about characterizing the problem space as we find it and foreshadowing some existing solutions we already use and some improvement proposals I want to make.

### Syntactical noise and readability

The alternation between literal strings and strings stored as character objects causes a lot syntactical noise.
This is why many of us embrace the interpolation offered by the [glue package](https://glue.tidyverse.org), which I'll use as a running comparator.

```{r}
library(glue)

path <- "path/to/a/thing"
message("Caching nycflights db at ", path)

# glue-ified
message(glue("Caching nycflights db at {path}"))

dim_f <- c(4, 7)
units <- "inch"
message("Saving ", dim_f[1], " x ", dim_f[2], " ", units, " image")

# glue-ified
message(glue("Saving {dim_f[1]} x {dim_f[2]} {units} image"))
```

`sprintf()` offers some hope of relief, because at least the single format string eliminates lots of quotes.

```{r}
name <- "thing"
replacement <- " by doohickey"
version <- "1.0.0"
message(sprintf("%s is deprecated%s. Deprecated in version '%s'.",
                name, replacement, version))

# glue-ified
message(glue(
  "{name} is deprecated{replacement}. Deprecated in version '{version}'."
))
```

I'd still rather write, read, and maintain the glue version.
You don't have to bounce between the format string and the arguments supplying values, doing tedious counting or substitution in your head.

Conclusion: string interpolation is critical for any function that makes messages.

### Literal quotes (or other inline style)

Often you want to surround an interpolated stored string with an actual quoting symbol, so it's clearly distinguished as, e.g., a variable or package name or a field.
This invites a very specific set of mistakes: not enough backslashes, backslashes in the wrong place, and unbalanced quotes.

<aside>
If available, you might use color and/or other styling (e.g., bold, underline) for this purpose.
</aside>

```{r}
FMT <- "FORMAT"
message("Using: \"", FMT, "\"")

element <- "ggplot2.thingy"
message("Theme element `", element, "` missing")

# glue-ified
message(glue("Using: \"{FMT}\""))
message(glue('Using: "{FMT}"'))   # some single / double quote games
message(glue("Theme element `{element}` missing"))
```

<aside>
Yes, you can play some single vs. double quote games, but it only gets you so far.
</aside>

Conclusion: Messages are not just "flat" strings.
You really need inline styles to convey all the necessary information.

Spoiler: this is the motivation for the cli package's [semantic CLI functionality](https://cli.r-lib.org/articles/semantic-cli.html), which has roots in the [`ui_*()` functions](https://usethis.r-lib.org/reference/ui.html) in usethis.

### Collapsing lists of things

It's pretty common that your message incorporates a collection of items
Here you see a lot of `paste()` and friends, used with the `collapse` argument.

```{r}
x <- structure(1, class = c("alfa", "bravo", "charlie"))
message("Don't know how to automatically pick scale for object of type ",
        paste(class(x), collapse = "/"), ". Defaulting to continuous.")

# glue-ified
message(glue("
 Don't know how to automatically pick scale for object of type \\
 {glue_collapse(class(x), sep = '/')}. Defaulting to continuous."            
))

msg <- c("'this'", "'that'")
message("`geom_smooth()` using ", paste0(msg, collapse = " and "))

# glue-ified
message(glue("`geom_smooth()` using {glue_collapse(msg, sep = ' and ')}"))
```

Conclusion: Often we want to inject an array of strings and collapse it with some attention to aesthetics, conventions, and grammar.

<aside>
The need to collapse lists often co-occurs with the need for inline style.
A closely related problem is allocating message parts (e.g. list items) to different lines, e.g. making bullet lists.
</aside>

The next example really starts to show multiple ergonomic problems interacting with each other, i.e. working over a list of interpolated strings and dealing with line breaks.

```{r}
orig_name <- c(a = "o1", b = "o2")
name <- c(a = "n1", b = "n2")
new_names <- c("b", "a")

message(
  "New names:\n",
  paste0(orig_name[new_names], " -> ", name[new_names], collapse = "\n")
)

# glue-ified
nms <- glue("{orig_name[new_names]} -> {name[new_names]}\n")
nms <- glue_collapse(nms, sep = "\n")
message(glue("New names:\n{nms}"))
```

We're also starting to see some challenges with glue, where you need to make several calls in sequence, to build up the message.
I've decided the explicit management of newlines (`\n`) is a strong indicator that there's some missing bit of tooling.

Let's go out with an extravaganza of interpolation, inline styling and collapsing!

```{r}
trained <- c(alfa = "apple", bravo = "banana")
message("Multiple formats matched: ",
        paste("\"", names(trained), "\"(", trained, ")", sep = "",
              collapse = ", "))

# glue-ified
x <- glue('"{names(trained)}"({trained})')
y <- glue_collapse(x, sep = ", ")
message(glue("Multiple formats matched: {y}"))
```




